Blueprint for efficient nuclear spin characterization with color center

### 2024-05-18 PRB - revision reply/main_revision.tex
last update?

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.2 distribution.
%   Version 4.2a of REVTeX, December 2014
%
%   Copyright (c) 2014 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.2
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[%
 reprint,
superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}
\usepackage{physics}
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{hyperref}% add hypertext capabilities
\usepackage{soul}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\newcommand{\vvv}[1]{\textcolor{blue}{#1}}
%\usepackage[showframe,%Uncomment any one of the following lines to test 
%%scale=0.7, marginratio={1:1, 2:3}, ignoreall,% default settings
%%text={7in,10in},centering,
%%margin=1.5in,
%%total={6.5in,8.75in}, top=1.2in, left=0.9in, includefoot,
%%height=10in,a5paper,hmargin={3cm,0.8in},
%]{geometry}
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%================= %THEORY
\section{Results}
\label{sec:theory}
\begin{figure*}%[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_1.pdf}
		\caption{Simulated DD signal with 64 $\pi$-pulses for the register reported in  \cite{abobeih2019atomic} consisting of 23 nuclear spins, assuming the electron spin is a. spin $S=1$ system b. spin $S=1/2$ system. Simulated Fisher information matrix for the case of the electron spin c. spin $S=1$ system d. spin $S=1/2$ system. The first 23 hyperfine parameters are $A_{zz}$ and the second 23 parameters are $A_{zx}$ of the nuclear spin register.}
		\label{fig:3}
	\end{center}
\end{figure*}
%%%%%%%%%%%%%%%Í%%%%=================
We consider a central spin system of \textit{non-interacting} nuclear spins $I=1/2$ coupled to a central electron spin $\boldsymbol{S}$.
The central electron spin is manipulated resonantly with microwave pulses, which transfer the population between the two sublevels denoted as $m_s=s_0$ and $m_s = s_1$.
These two spin sublevels are separated due to the Zeeman effect and/or zero-field splitting, forming a two-level subsystem with an energy splitting of $\omega_a$.
The Hamiltonian of the system in the secular approximation and in the rotating frame of the applied microwave $\omega_{mw}$ can be written as:
\begin{align}
	H = \Delta S_z + \sum_k (\omega_L^{(k)}  + A_{zz}^{(k)} S_z) I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
	\label{eq:H}
\end{align}
Where $\Delta = \omega_a - \omega_{mw}$ is the detuning, $\omega_L^{(k)} =\gamma_n^{(k)} B$ is the nuclear Rabi frequency with $\gamma_n^{(k)}$ being the nuclear spin gyromagnetic ratio and $B$ the external magnetic field, and $A_{zz}$ and $A_{zx}$ are the parallel and perpendicular secular components of the hyperfine tensor, respectively.
The Hamiltonian is diagonal in the electron spin subspace, thus the Hamiltonian for the nuclear spins could be rewritten in the electron spin subdomains $s_i$ ($i=0,1$).
The nuclear spin Hamiltonian could be solved for eigenenergies, which determine the precession frequencies of $\omega_i = \sqrt{(\omega_L + s_i A_{zz})^2+ (s_i A_{zx})^2}$, $i=0,1$ along the eigenvector axes $\vec{n}_i =(\frac{s_i A{zx}}{\omega_i}, 0, \frac{\omega_L + s_i A_{zz}} {\omega_i})$.
It is assumed that the pulse duration $t_p$ is short enough ($\omega_L t_p \ll 1$) that the nuclear spin dynamics during this time is negligible.
For completeness, we start our analysis with the simplest characterization sequence, which is the electron spin Ramsey sequence (free induction decay of the electron).
The Ramsey signal can be obtained as follows:
\begin{gather}
	\expval{\sigma_z}_{Ram} = \cos(\Delta \tau) \prod_{j=1}^{n} \Big[ \cos(\frac{\omega_0 \tau}{2}) \cos(\frac{\omega_1 \tau}{2})+ \notag \\
	(\vec{n_0}\cdot \vec{n_1})\sin(\frac{\omega_0 \tau}{2}) \sin(\frac{\omega_1 \tau}{2}) \Big]^{(j)}
	%\textcolor{gray}{e^{-(\frac{\tau}{T_2^\ast})^m}}
\end{gather}	
\st{The exponential decay is introduced to note the Ramsey signal decays with the electron spin $T_2^\ast$ since the formalism does not account for any relaxation processes.}  \textcolor{blue}{We note the expressions do not account for any relaxation processes.} The Ramsey sequence is sensitive to detuning since the electron spin is not refocused \textcolor{blue}{and it decays with the electron spin $T_2^\ast$}; however, it is possible to sense nuclear spins with vanishing perpendicular hyperfine coupling.\\

@@

@@

>>>>

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.2 distribution.
%   Version 4.2a of REVTeX, December 2014
%
%   Copyright (c) 2014 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.2
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[%
 reprint,
superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{physics}
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{hyperref}% add hypertext capabilities
\usepackage{soul}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\newcommand{\vvv}[1]{\textcolor{blue}{#1}}
\newcommand{\kb}[2]{\vert #1 \rangle \langle #2 \vert}
\newcommand{\caver}[1]{\overline{#1}}
\setlength{\jot}{8pt}
%\newcommand{\abs}[1]{\vert #1 \vert}
\let\oldsubref\subref
%\renewcommand{\subref}[1]{(\oldsubref{#1})}
\renewcommand{\vec}[1]{\overrightarrow{#1}}
\newcommand{\tq}[1]{\textquote{#1}}
%\usepackage[showframe,%Uncomment any one of the following lines to test 
%%scale=0.7, marginratio={1:1, 2:3}, ignoreall,% default settings
%%text={7in,10in},centering,
%%margin=1.5in,
%%total={6.5in,8.75in}, top=1.2in, left=0.9in, includefoot,
%%height=10in,a5paper,hmargin={3cm,0.8in},
%]{geometry}
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%================= %THEORY
\section{Results}
\label{sec:theory}
\begin{figure*}%[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_1.pdf}
		\caption{Simulated DD signal with 64 $\pi$-pulses for the register reported in  \cite{abobeih2019atomic} consisting of 23 nuclear spins, assuming the electron spin is a. spin $S=1$ system b. spin $S=1/2$ system. Simulated Fisher information matrix for the case of the electron spin c. spin $S=1$ system d. spin $S=1/2$ system. The first 23 hyperfine parameters are $A_{zz}$ and the second 23 parameters are $A_{zx}$ of the nuclear spin register.}
		\label{fig:3}
	\end{center}
\end{figure*}
%%%%%%%%%%%%%%%Í%%%%=================
We consider a central spin system of \textit{non-interacting} nuclear spins $I=1/2$ coupled to a central electron spin $\boldsymbol{S}$.
The central electron spin is manipulated resonantly with microwave pulses, which transfer the population between the two sublevels denoted as $m_s=s_0$ and $m_s = s_1$.
These two spin sublevels are separated due to the Zeeman effect and/or zero-field splitting, forming a two-level subsystem with an energy splitting of $\omega_a$.
The Hamiltonian of the system in the secular approximation and in the rotating frame of the applied microwave $\omega_{mw}$ can be written as \textcolor{blue}{(derivation in \ref{app:Hamiltonian})}:
\begin{align}
	H = \Delta S_z + \sum_k (\omega_L^{(k)}  + A_{zz}^{(k)} S_z) I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
	\label{eq:H}
\end{align}
Where $\Delta = \omega_a - \omega_{mw}$ is the detuning, $\omega_L^{(k)} =\gamma_n^{(k)} B$ is the nuclear Rabi frequency with $\gamma_n^{(k)}$ being the nuclear spin gyromagnetic ratio and $B$ the external magnetic field, and $A_{zz}$ and $A_{zx}$ are the parallel and perpendicular secular components of the hyperfine tensor, respectively.
The Hamiltonian is diagonal in the electron spin subspace, thus the Hamiltonian for the nuclear spins could be rewritten in the electron spin subdomains $s_i$ ($i=0,1$).
The nuclear spin Hamiltonian could be solved for eigenenergies, which determine the precession frequencies of $\omega_i = \sqrt{(\omega_L + s_i A_{zz})^2+ (s_i A_{zx})^2}$, $i=0,1$ along the eigenvector axes $\vec{n}_i =(\frac{s_i A{zx}}{\omega_i}, 0, \frac{\omega_L + s_i A_{zz}} {\omega_i})$.
It is assumed that the pulse duration $t_p$ is short enough ($\omega_L t_p \ll 1$) that the nuclear spin dynamics during this time is negligible.
For completeness, we start our analysis with the simplest characterization sequence, which is the electron spin Ramsey sequence (free induction decay of the electron).
The Ramsey signal can be obtained as follows:
\begin{gather}
	\expval{\sigma_z}_{Ram} = \cos(\Delta \tau) \prod_{j=1}^{n} \Big[ \cos(\frac{\omega_0 \tau}{2}) \cos(\frac{\omega_1 \tau}{2})+ \notag \\
	(\vec{n_0}\cdot \vec{n_1})\sin(\frac{\omega_0 \tau}{2}) \sin(\frac{\omega_1 \tau}{2}) \Big]^{(j)}
	%\textcolor{gray}{e^{-(\frac{\tau}{T_2^\ast})^m}}
\end{gather}	
\st{The exponential decay is introduced to note the Ramsey signal decays with the electron spin $T_2^\ast$ since the formalism does not account for any relaxation processes.}  \textcolor{blue}{We note the expressions do not account for any relaxation processes.} The Ramsey sequence is sensitive to detuning since the electron spin is not refocused \textcolor{blue}{and it decays with the electron spin $T_2^\ast$}; however, it is possible to sense nuclear spins with vanishing perpendicular hyperfine coupling.\\

@@

\section{Hamiltonian}
\label{app:Hamiltonian}
The spin Hamiltonian for many color centers can be written as \cite{schweiger2001principles}:
\begin{gather}
	H = \mathbf{D}^\intercal \mathbf{S} \mathbf{D} + \beta_e \mathbf{B}^\intercal \mathbf{g} \mathbf{S} + \beta_n \sum_{k=1}^m g_n^{(k)}  \mathbf{B}^\intercal \mathbf{I}^{(k)} \notag \\
	+ \sum_{I_k>1/2} (\mathbf{I}^{\intercal} \mathbf{P} \mathbf{I})^{(k)} + \sum_{k=1}^m g_n^{(k)} (\mathbf{S}^\intercal \mathbf{A} \mathbf{I})^{(k)}
\end{gather}
Where $\mathbf{S}$ and $\mathbf{I}^{(k)}$ are electron and nuclear spin vector operators, $\mathbf{D}$ is zero field splitting tensor, $\mathbf{B}$ is magnetic field vector, $\mathbf{P}$ is nuclear quadrupole tensor, $\mathbf{A}$ is the hyperfine interaction tensor, $\beta_e = \frac{e \hbar}{2m_e}$ and $\beta_n = \frac{-e \hbar}{2m_n}$ are electron and nuclear Bohr magneton. Here, we neglected  nuclear-nuclear interactions. Also, due to axial symmetry, the zero field splitting can be simplified. We assume the magnetic field is only along the symmetry axis, which we denote as z direction. Moreover, the secular terms ($S_zI_x, S_zI_y, S_zI_z$) are just a shift of the energy level that are relevant up to first order, while the non-secular terms ($S_xI_x, S_yI_x, S_zI_z, S_xI_y, S_yI_y, S_zI_y$) mix the electron spin levels, which are largely detuned due to the large zero field splitting. Hence, the non-secular terms of Hamiltonian are neglected. Furthermore, for each nuclear spin, one can go to the xy frame such that $A_{zy}$ term is zero. All in all, the spin Hamiltonian with mentioned considerations can be simplified to:
\begin{gather}
	H = D_z S_z^2 + \gamma_e B S_z + \gamma_n B I_z^{(k)} + P_z^{(k)} I_z^{2(k)} \notag \\
	+ A_{zz}^{(k)} S_z I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
\end{gather}
The summation over k is omitted due to convenience!
The coefficient $\gamma_e$ and $\gamma_n$ are electron and nuclear gyromagnetic ratios. The coefficients' values for the case of NV center in diamond are $D_z= 2.87$ GHz, $\gamma_e= -2.802$ MHz/G, $\gamma_n^{(^{13}C)}= 1.07$ kHz/G, $\gamma_n^{(^{14}N)}= 0.31$ kHz/G, $P_z^{(^{14}N)}= -5.04$ MHz. \\

The electron spin in general can take any value, however resonant microwave pulses are applied only for two of the sublevels. Hence, the electron spin is effectively a two-level system. The electron spin occupies two of its sublevels that are denoted as $\ket{0} = \ket{m_s=s_0}$ and $\ket{1} = \ket{m_s=s_1}$; in other words:
\begin{gather}
	S_z= \begin{pmatrix} s_1 & 0 \\ 0 & s_0\end{pmatrix}
\end{gather}
One can show that $S_z^2 = (s_1+s_0)S_z-s_1s_0 \mathds{1}$. 
We define the transition frequency between two spin sub-levels of the electron spin as $\omega_a = D_z (s_1+s_0)+\gamma_e B $. Moreover, we note that since we consider nuclei with spin of one half in the system, the nuclear quadrupole term vanishes. The Hamiltonian can be written in terms of Pauli matrices:
\begin{align}
	H =\omega_aS_z+ \gamma_n B I_z^{(k)} + A_{zz}^{(k)} S_z I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
\end{align}

The Hamiltonian presented so far does not include the microwave drive field. Indeed, the microwave pulses are assumed to be instantaneous, but considering the drive field, the Hamiltonian  should be written in the rotating frame and assume rotating wave approximation. This frame should be respected for the rest of sequence. Since the rotating frame unitary operator commutes with the $\sigma_z$, the expectation value of this operator in the rotating frame is the same as in the lab frame. Eventually, free evolution Hamiltonian in the rotating frame is written as follows:
\begin{align}
	H = \Delta S_z + \sum_k (\omega_L^{(k)}  + A_{zz}^{(k)} S_z) I_z^{(k)} + A_{zx}^{(k)} S_z I_x^{(k)}
\end{align}
Where $\Delta = \omega_a - \omega_{mw}$ is the detuning.

@@

\bibitem [{\citenamefont {Schweiger}\ and\ \citenamefont
  	{Jeschke}(2001)}]{schweiger2001principles}%
  		{Schweiger}}\ and\ \bibinfo {author} {\bibfnamefont {G.}~\bibnamefont
  		{Jeschke}},\ }\bibfield  {title} {\bibinfo {title} {Principles of pulse electron paramagnetic resonance} }\href@noop
  {} {\bibfield  {book} {\bibinfo  {book} {Oxford university press}\ }
  	(\bibinfo {year} {2001})}\BibitemShut {NoStop}%

### 2024-05-18 PRB - revision reply/reply.tex
last update?

	4. \textcolor{red}{We need to either add a reference or a small paragrah saying why Secular is applied and a $U_{int}$ to rotation frame, now it is too rude answer}The Hamiltonian in Eq. 1 is very well-understood in this community [ref here]. The Hamiltonian conventionally is written in the rotating frame of applied microwave and consist of Electron and nuclear spin Zeeman energy and dipole interaction between them. As it is mentioned in the text, the secular approximation is considered so terms like $S_x, S_y$ do not appear. Traditionally, the nuclear spins' frame are rotated such that $I_y$ term do not appear. This Hamiltonian applied to any two-level system, e.g., this could be any two spin sublevels in Silicon Carbide (S=3/2)\\

@@

	6. \textcolor{red}{Initial state should be initialised into superposition of two eigenstates, s0,s1, which can be any, right?}The initial state for the Ramsey sequence can be any of the two sub-levels of spin system. Indeed, the grey exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion. We have added this correction to the main text.
	\textcolor{blue}{((..initial state is XXX..))}.\\

@@

	9. In this work, we address nuclear spins only with microwave driving of the electron spin (\textcolor{red}{did we mention this in the text?}). Applying another radio frequency (RF) driving adds another experimental challenge to the problem and goes beyond the scope of this work. In this case, RF spectroscopy to find and drive individual nuclear spins is required, which can add the required time to image nuclear spins, which was successfully demonstrated in various works for spin 1, 3/2,  \cite{DDRF, ENDOR, SSRV2}. We believe that it will as well work well for spin $S=1/2$, as it probes the transition frequency of the nuclear spin in $m_s=1/2$ or $m_s=-1/2$ subdomains, making it sensitive to the first order of the $A_{zz}$ coupling. Comparison of various RF assisted protocols goes beyond the scope of the current work and requires further investigation.\\

@@

	1. Characterization of spin-1 system with both DD and DDESEEM works similarly well. Indeed, it is important to observe the diagonal and side-diagonals are distinguishable from in the Fisher Information matrix, which in both Fig. 2c and 3c are clear. However, the color scale is levelled such that in each sequence, the perfomance of spin 1/2 can be compared with spin 1. Fig. 3c, 3d are Fisher Information matrix simulated for 5pESEEM. For clarity we added this point in the caption.\\

>>>>

	4. \textcolor{brown}{We thank the referee to point out that this Hamiltonian needs further explanation. We added an Appendix to explain the derivation of this Hamiltonian.} The Hamiltonian conventionally is written in the rotating frame of applied microwave and consist of Electron and nuclear spin Zeeman energy and dipole interaction between them. As it is mentioned in the text, the secular approximation is considered so terms like $S_x, S_y$ do not appear. Traditionally, the nuclear spins' frame are rotated such that $I_y$ term do not appear. This Hamiltonian applied to any two-level system, e.g., this could be any two spin sublevels in Silicon Carbide (S=3/2)\\

@@

	6. \textcolor{red}{Initial state should be initialised into superposition of two eigenstates, s0,s1, which can be any, right?}
	\textcolor{brown}{No, the initial state is always one of s0 or s1, with the first pi-half pulse, the superposition is created. Or one can say I know how to generate an initial superposition state, then the first pi-half pulse is not needed.}
	The initial state for the Ramsey sequence can be any of the two sub-levels of spin system. Indeed, the grey exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion. We have added this correction to the main text.
	\textcolor{blue}{((..initial state is XXX..))}.\\

@@

	9. In this work, we address nuclear spins only with microwave driving of the electron spin as we mentioned in the text:\\
	\textcolor{brown}{((While double resonance methods hold large promise for nuclear spin spectroscopy, they require additional experimental hardware, and often are not available.))}
	Applying another radio frequency (RF) driving adds another experimental challenge to the problem and goes beyond the scope of this work. In this case, RF spectroscopy to find and drive individual nuclear spins is required, which can add the required time to image nuclear spins, which was successfully demonstrated in various works for spin 1, 3/2,  \cite{DDRF, ENDOR, SSRV2}. We believe that it will as well work well for spin $S=1/2$, as it probes the transition frequency of the nuclear spin in $m_s=1/2$ or $m_s=-1/2$ subdomains, making it sensitive to the first order of the $A_{zz}$ coupling. Comparison of various RF assisted protocols goes beyond the scope of the current work and requires further investigation.\\

@@

	1. Characterization of spin-1 system with both DD and DDESEEM works similarly well. Indeed, it is important to observe the diagonal and side-diagonals, which are clearly distinguishable in the Fisher Information matrix in both Fig. 2c and 3c. However, the color scale is levelled such that in each sequence, the perfomance of spin 1/2 can be compared with spin 1. Fig. 3c, 3d are Fisher Information matrix simulated for 5pESEEM. For clarity we added this point in the caption.\\
